ODE and Diffusion Models of Tumor Growth

### 2024-03-14 paper.tex
5.5

\documentclass{article}
\usepackage{graphicx}

\title{ODE and Diffusion Models of Tumor Growth}
\author{Daniel Henricks}

@@

Researchers have been studying cancer and the patterns behind tumor growth for thousands of years. Hippocrates
was accredited with possibly being the first to describe a tumor, and tumors were mainly described by their color
and their response to touch. However, ancient physicians did not separate benign and malignant tumors, but they
acknowledged the tumor's ability to infect nearby tissue. The application of mathematical models to tumor growth
did not start to gain traction until much later, however. In the early 20th century, one of the most famous uses of
the diffusion model was first developed: Hill released a paper discussing how the movement of lactic acid
and oyxgen throughout the body could be shaped into a diffusion problem (Hill, 1928). This paper eventually led to
Burton's paper which modeled the growth of tumors using diffusion, applying many of the mathematical tools employed
by Hill some years before (Burton, 1966). This paper will present a brief overview of the mathematics used in these
papers and will recreate some of the graphics produced as well.

@@

Hill then combined both equations to come to a conclusion on how lactic acid and oxygen would interact in a muscle.
The conclusions made from this section of the paper were essential for future models of diffusion. First, it was
assumed that the thickness $b > x'$ so none of the oxygen was able to penetrate through the entire system. From above, 
the concentrations of oxygen and lactic acid are $y$ and $y'$ respectively. Then the following graphic describes the system:

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9\textwidth]{graphics/image.png}
    \caption{From 0 to 5 and 20 to 25, the concentration $y$ is equal to the steady state $y_0$. From 5 to 10,
    the concentration $y$ is decreasing until it reaches 0 at 10 (in green). From 10 to 15, the lactic acid concentration
    $y'$ (in blue) is increasing to its maximum and then decreases to 0 at 15. From 15 to 20, the oxygen concentration $y_0$ increases
    until reaching the steady state (red) at x = 20.}
    \label{fig:image}
\end{figure}

>>>>

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\title{ODE and Diffusion Models of Tumor Growth}
\author{Daniel Henricks}

@@

Researchers have been studying cancer and the patterns behind tumor growth for thousands of years. Hippocrates
was accredited with possibly being the first to describe a tumor, and tumors were mainly described by their color
and their response to touch. However, ancient physicians did not separate benign and malignant tumors, but they
acknowledged the tumor's ability to infect nearby tissue. The application of mathematical models to tumor growth
did not start to gain traction until much later, however. In the early 20th century, one of the most famous uses of
the diffusion model was first developed: Hill released a paper discussing how the movement of lactic acid
and oyxgen throughout the body could be shaped into a diffusion problem (Hill, 1928). This paper eventually led to
Greenspan's paper which modeled the growth of tumors using diffusion, applying many of the mathematical tools employed
by Hill some years before (Greenspan, 1972). This paper will present a brief overview of the mathematics used in these
papers and will recreate some of the graphics produced as well.

@@

Hill then combined both equations to come to a conclusion on how lactic acid and oxygen would interact in a muscle.

The conclusions made from this section of the paper were essential for future models of diffusion. First, it was
assumed that the thickness $b > x'$ so none of the oxygen was able to penetrate through the entire system. From above,
the concentrations of oxygen and lactic acid are $y$ and $y'$ respectively. The plane sheet of tissue will have oxygen
at a partial pressure of $y_0$ from the left and an impenetrable wall on the right. Also let $a$ be the rate of usage
(in cubic centimeters of tissue) of oxygen by the tissue.
Then the following graphic describes the system:

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{graphics/image.png}
    \caption{From 0 to 5 and 20 to 25, the concentration $y$ is equal to the steady state $y_0$. From 5 to 10,
        the concentration $y$ is decreasing until it reaches 0 at 10 (in green). From 10 to 15, the lactic acid concentration
        $y'$ (in blue) is increasing to its maximum and then decreases to 0 at 15. From 15 to 20, the oxygen concentration $y_0$ increases
        until reaching the steady state (red) at x = 20. Note that the $x$ and $y$ values for this graphic are arbitrary.}
    \label{fig:image}
\end{figure}

The diffusion equations that satisfy this model are easily seen from the graphic and discussion above:

$-\alpha = k' \frac{d^2y'}{dx^2} $(the source of lactic acid on the right)

$a = k \frac{d^2y}{dx^2}$ (the oxygen from the left.)

\vspace*{0.25cm}
The boundary conditions for these PDEs are:
\begin{enumerate}
    \item[(a)] $y = y_0$ when x is to the left of the source of oxygen diffusion.
    \item[(b)] $\frac{dy'}{dx} = 0$ for $x = b$ since no diffusion of oxygen occurs here.
    \item[(c)] The concentration of lactic acid must become zero at the same point as the concentration of oxygen becomes zero.
\end{enumerate}

At the point where (c) occurs, the diffusion of oxygen from the left must be equal to the diffusion of lactic acid from the right.
Mathematically, this can be expressed as the equation

\begin{equation}
    -k \frac{dy}{dx} = k' \frac{dy}{dx}
\end{equation}

Expressed as a ratio and simplified, this means that:

\begin{equation}
    \frac{-k \frac{dy}{dx}}{k' \frac{dy}{dx}} = \frac{-k\frac{ax}{k}}{k'\frac{-\alpha x}{k'}} = \frac{a}{\alpha}
\end{equation}

This equation states that in this steady state, it takes $\frac{a}{\alpha} cm^3$ of oxgyen to remove 1 gram of lactic acid. Hill continues
the paper by discussing other possible steady states that could be used to investigate this model. However, they are not of interest for
this paper.

\section{Greenspan's paper: Models for the Growth of a Solid Tumor by Diffusion}

Greenspan's paper was released almost 50 years after the inital paper by Hill was authored. Since then, many mathematical models
had been employed to try to model processes within the body. A few years prior to the release of Greenspan's paper, Burton (1966)
authored one of the most famous papers on the growth of tumors as a problem of diffusion. Greenspan expanded on many of the models
developed by Burton.

The primary goal of Greenspan was to "infer the chemical source of growth inhibition from the most easily obtained data...the outer radius
of the nodule as a function of time." The conclusions were also described "with as little mathematics as possible" to try to attract a more
general audience, unlike the previous papers of the time.
